\thetitle

### 2016-02-12 internal_homs.tex
asterisk up

\section{Elementary properties of E-systems}\label{sec:esys_props}

The theory of E-systems could be seen as a sort of category theory with only
projections and their sections. We elaborate on this point of view by defining
the pre-category of internal morphisms, which Lawvere might have called the
category of terms. 
In the pre-category of internal morphisms over a context $\Gamma$,
which we define in \autoref{subsec:im_cat},
a morphism $f:A\to B$ is a term of the constant family $W_A(B)$ over 
$\ctxext{\Gamma}{A}$. Rather than directly defining composition, we will define
for every $f:A\to B$ in context $\Gamma$, a pre-composition operation
$f_\ast$, which is going to be an E-homomorphism from 
$\cat{F}/\ctxext{\Gamma}{B}$ to $\cat{F}/\ctxext{\Gamma}{A}$. This allows us
to consider diagrams of the following form:
\begin{equation*}
\begin{tikzcd}
P \arrow[d,->>] \arrow[r,"g"] & Q \arrow[d,->>] \\
A \arrow[r,swap,"f"] & B
\end{tikzcd}
\end{equation*}
where we consider the morphism $g$ over $f$ to be an internal morphism from 
$P$ to $f_\ast(Q)$ in context $\ctxext{\Gamma}{A}$, capturing the
essence of being a morphism from the family $P$ over $A$ to the family
$Q$ over $B$, reindexed by $f$. 
Whenever we draw such diagrams, a double headed arrow will always
refer to a family, and an ordinary arrow will refer to an internal morphism,
possibly over another internal morphism.

@@

The pre-category of internal morphisms forms a pre-category of which the families form
a subpre-category. This, together with the pre-composition operation described
earlier, gives rise to the auxiliary notion of E'-systems, which we define
in \autoref{sec:E'}. The thread running through this whole section is proving
that the notions of E-system and E'-system are the same. In \autoref{sec:E'}
we show directly that any E'-system is an E-system. We end this section by
showing that the converse also holds.

@@

\begin{defn}\label{defn:E'sys}
An \define{E'-system} consists of a pre-category $\cat{C}$ and a subpre-category $\cat{F}$ with
the same objects as $\cat{C}$, such that
\begin{enumerate}
\item $\cat{F}$ has a terminal object.
\item For any $f:\Delta\to\Gamma$ in $\cat{C}$ and any $A\in\cat{F}/\Gamma$, a
functorial choice of a pullback square 
\begin{equation*}
\begin{tikzcd}[column sep=large]
\ctxext{\Delta}{f_\ast(A)} \arrow[r,"{\pi_2(f,A)}"] \arrow[d,fib,swap,"{f_\ast(A)}"] & \ctxext{\Gamma}{A} \arrow[d,fib,"A"] \\
\Delta \arrow[r,"f"] & \Gamma
\end{tikzcd}
\end{equation*}
such that $f_\ast(A)\in\cat{F}/\Delta$. Explicitly, the functoriality
requirement is that
\begin{enumerate}
\item For any $f:\Delta\to\Gamma$, one has 
$f_\ast(\catid{\Gamma})\jdeq\catid{\Delta}$ and
\begin{equation*}
\pi_2(f,\catid{\Gamma})\jdeq f.
\end{equation*}
\item \label{defn:E'sys:b} For any $A\in\cat{F}/\Gamma$, one has $(\catid{\Gamma})_\ast(A)\jdeq A$ and
\begin{equation*}
\pi_2(\catid{\Gamma},A)\jdeq\catid{\ctxext{\Gamma}{A}}
\end{equation*}
\item \label{defn:E'sys:c} For any $f:\Delta\to\Gamma$, $g:\greek{E}\to\Delta$ and $A\in\cat{F}/\Gamma$, one
has $(f\circ g)_\ast(A)\jdeq g_\ast(f_\ast(A))$ and 
\begin{equation*}
\pi_2(f\circ g,A)\jdeq \pi_2(f,A)\circ\pi_2(g,f_\ast(A))
\end{equation*}
\item For any $P\in\cat{F}/\ctxext{\Gamma}{A}$ and $f:\Delta\to\Gamma$, one has
$f_\ast(P)\jdeq\pi_2(f,A)_\ast(P)$ and 
\begin{equation*}
\pi_2(f,\ctxext{A}{P}) \jdeq \pi_2(\pi_2(f,A),P)
\end{equation*}
\end{enumerate}
\end{enumerate}
For any $f:\Delta\to\Gamma$ we shall write $f_\ast$ for the induced functor
$\cat{F}/\Gamma\to\cat{F}/\Delta$. 
\end{defn}

\begin{rmk}
Without a terminal object in $\cat{F}$, the pre-category $\cat{C}$ does not correspond
directly to anything in an E-system.
\end{rmk}

@@

\begin{lem}\label{lem:pb_selfdistributive}
Let $\mathbb{E}$ be an E'-system. Then $f_\ast$ can be extended to a functor from
$\cat{C}_\cat{F}/\Gamma$ to $\cat{C}_\cat{F}/\Delta$, for each $f:\Delta\to\Gamma$, and moreover the diagram
\begin{equation*}
\begin{tikzcd}[column sep=large]
\cat{C}_\cat{F}/\ctxext{\Gamma}{B} \arrow[r,"{f_\ast/B}"] \arrow[d,swap,"{g_\ast}"]
& \cat{C}_\cat{F}/\ctxext{\Delta}{f_\ast(B)} \arrow[d,"{(f_\ast(g))_\ast}"] \\
\cat{C}_\cat{F}/\ctxext{\Gamma}{A} \arrow[r,swap,"{f_\ast/A}"]
& \cat{C}_\cat{F}/\ctxext{\Delta}{f_\ast(A)}
\end{tikzcd}
\end{equation*}
commutes for every $g:A\to B$ in $\cat{C}_\cat{F}/\Gamma$.
\end{lem}

\begin{proof}
It is straightforward to use the universal property of the pullback to extend
$f_\ast$ to a functor from $\cat{C}_\cat{F}/\Gamma$ to $\cat{C}_\cat{F}/\Delta$.
To see that the diagram commutes, we calculate
\begin{align*}
f_\ast/A\circ g_\ast
& \jdeq
\pi_2(f,A)_\ast\circ g_\ast \\
& \jdeq
(g\circ\pi_2(f,A))_\ast \\
& \jdeq
(\pi_2(f,B)\circ f_\ast(g))_\ast \\
& \jdeq
(f_\ast(g))_\ast\circ\pi_2(f,B)_\ast \\
& \jdeq
(f_\ast(g))_\ast\circ f_\ast/B.\qedhere
\end{align*}
\end{proof}

\begin{thm}
An E-system is the same thing as an E'-system.
\end{thm}

\begin{proof}
We define for any $A\in\cat{F}/\Gamma$ the set
\begin{equation*}
T(A)\defeq\{x:\Gamma\to\ctxext{\Gamma}{A}\mid A\circ x\jdeq\catid{\Gamma}\}.
\end{equation*}
Since each functor $f_\ast$ is extended to all of $\cat{C}_{\cat{F}}/\Gamma$,
it follows immediately that those are functors with term structure.

We define for any $A\in\cat{F}/\Gamma$, the functor $W_A\defeq A_\ast:\cat{F}/\Gamma\to\cat{F}/\ctxext{\Gamma}{A}$.
Likewise, we define for any $x\in T(A)$, the functor $S_x\defeq x_\ast:\cat{F}/\ctxext{\Gamma}{A}\to\cat{F}/\Gamma$.
We also define $\tfid{A}:T(W_A(A))$ by the universal property of pullbacks:
\begin{equation*}
\begin{tikzcd}[column sep=large]
\ctxext{\Gamma}{A} \arrow[drr,bend left=15,equals] \arrow[ddr,bend right=15,equals] \arrow[dr,densely dotted,near end,"{\tfid{A}}"] \\
& \ctxext{{\Gamma}{A}}{W_A(A)} \arrow[r,swap,"{\pi_2(A,A)}"] \arrow[d,fib] & \ctxext{\Gamma}{A} \arrow[d,fib] \\
& \ctxext{\Gamma}{A} \arrow[r,swap,"A"] & \Gamma
\end{tikzcd}
\end{equation*}
As an immediate consequence of \autoref{lem:pb_selfdistributive}, we get that
each functor $W_A$ and $S_x$ is both a weakening functor and a substitution functor.
Since each $W_A$ and $S_x$ preserves identity morphisms, it also preserves the
terms $\tfid{A}$.

It remains to verify the rest of the conditions of E-systems.
\begin{itemize}
\item To show that weakened families are constant, note that
\begin{equation*}
S_x\circ W_A\jdeq x_\ast\circ A_\ast\jdeq (A\circ x)_\ast\jdeq (\catid{\Gamma})_\ast\jdeq \catid{\cat{C}_{\cat{F}}/\Gamma}.
\end{equation*}
\item The identity terms are neutral for pre-composition:
\begin{equation*}
S_{\tfid{A}}\circ W_A/A \jdeq S_{\tfid{A}}\circ \pi_2(A,A)_\ast \jdeq (\pi_2(A,A)\circ\tfid{A})_\ast \jdeq
(\catid{\ctxext{\Gamma}{A}})_\ast \jdeq \catid{\cat{C}_{\cat{F}}/\ctxext{\Gamma}{A}} 
\end{equation*}
\item The identity terms behave like identity functions: by the universal property,
$S_x(\tfid{A})$ is the unique section of $A$ such that the square
\begin{equation*}
\begin{tikzcd}[column sep=6em]
\Gamma \arrow[d,swap,"{S_x(\tfid{A})}"] \arrow[r,"{\pi_2(x,\catid{\ctxext{\Gamma}{A}})}"] & \ctxext{\Gamma}{A} \arrow[d,"{\tfid{A}}"] \\
\ctxext{\Gamma}{A} \arrow[r,swap,"{\pi_2(x,W_A(A))}"] & \ctxext{{\Gamma}{A}}{W_A(A)}
\end{tikzcd}
\end{equation*}
commutes. Thus, it suffices to show that this square also commutes with $x$ in the place of
$S_x(\tfid{A})$. Note that $\pi_2(x,\catid{\ctxext{\Gamma}{A}})\jdeq x$. 
Since $\ctxext{{\Gamma}{A}}{W_A(A)}$ is itself a pullback, it suffices
and it is straightforward to verify the equalities
\begin{align*}
W_A(A)\circ\pi_2(x,W_A(A))\circ x & \jdeq W_A(A)\circ\idtm{A}\circ x \\
\pi_2(A,A)\circ\pi_2(x,W_A(A))\circ x & \jdeq \pi_2(A,A)\circ\idtm{A}\circ x.\qedhere
\end{align*}
\end{itemize}
\end{proof}

\begin{comment}
We then demonstrate how to define a category internal to an E-system. The main
point of this lies mainly in demonstrating how the operations of E-systems can
be used to define a category, because it is no surprise that a category can be
defined in type theory. Nevertheless, it is noteworthy that we need no further 
conditions on E-systems.
This is unlike the situation in category theory, where we need
to assume that the ambient category has finite limits in order to say what an
internal category is. In fact, it is possible to define an E-system internal
to an E-system, just using the algebraic structure of an E-system, and actually
defining an internal category was the hardest part of that definition.
\end{comment}

@@

\begin{defn}
Let $A,B\in\cat{F}/\Gamma$.
For any $f\in\thom{A}{B}$ we define the \define{pre-composition} E-homomorphism
\begin{equation*}
f_\ast \defeq S_f\circ W_A/B : \cat{F}/\ctxext{\Gamma}{B}\to\cat{F}/\ctxext{\Gamma}{A}.
\end{equation*}
Furthermore, for $g\in\thom{B}{C}$ we may consider $W_B(C)\in\cat{F}/\ctxext{\Gamma}{B}$ as a morphism from $W_B(C)$ to $\catid{\ctxext{\Gamma}{B}}$. Therefore, $f_\ast$ acts on the terms of
$W_B(C)$, so it makes sense to define $g\circ f\defeq f_\ast(g)$. In fact, it makes
sense to let $(\blank)\circ f$ be the infix notation for $f_\ast$. 
\end{defn}

@@

\begin{rmk}
Note that \autoref{f:SidfWff_id} of \autoref{defn:esystems} asserts precisely
that ${(\tfid{A})}_\ast\jdeq \catid{\cat{F}/\ctxext{\Gamma}{A}}$ for any $A\in\cat{F}/\Gamma$. In
particular, it follows that $g\circ\tfid{A}\jdeq g$ for any $g\in\thom{A}{B}$
\end{rmk}

\begin{lem}\label{lem:compcomp}
For any $f\in\thom{A}{B}$ and $g\in\thom{B}{C}$ we have $f_\ast\circ g_\ast\jdeq (g\circ f)_\ast$. 
\end{lem}

\begin{proof}
\begin{align*}
f_\ast\circ g_\ast & \jdeq S_f\circ (W_A/B) \circ S_g \circ (W_B/C)\\
& \jdeq S_f\circ S_{W_A(g)}\circ(W_A/\ctxext{B}{W_B(C)})\circ (W_B/C) \\
& \jdeq S_{S_f(W_A(g))}\circ (S_f/W_A(W_B(C)))\circ (W_A/\ctxext{B}{W_B(C)}) \circ W_B/C \\
& \jdeq S_{S_f(W_A(g))}\circ ((S_f\circ (W_A/B) \circ W_B)/C) \\
& \jdeq S_{S_f(W_A(g))}\circ ((S_f\circ W_{W_A(B)}\circ W_A)/C) \\
& \jdeq S_{S_f(W_A(g))}\circ W_A/C \\
& \jdeq (g\circ f)_\ast.\qedhere
\end{align*}
\end{proof}

\begin{thm}
For each object $\Gamma$ of an E-system $\mathbb{E}$, we have a pre-category
$\mathbb{C}_I(\mathbb{E},\Gamma)$ of internal morphisms of $\mathbb{E}$,
consisting of the same objects as $\cat{F}/\Gamma$ and the hom-sets 
$\thom{A}{B}$. 
\end{thm}

\begin{proof}
The fact that composition is associative is a direct corollary of
\autoref{lem:compcomp}. The axiom $(\tfid{A})_\ast\jdeq\catid{\ctxext{\Gamma}{A}}$
implies that the identity morphisms satisfy the right identity law. It remains
to show that $\tfid{B}\circ f\jdeq f$. This is a simple calculation:
\begin{equation*}
\tfid{B}\circ f 
  \jdeq
\subst{f}{\ctxwk{A}{\tfid{B}}}
  \jdeq
\subst{f}{\tfid{\ctxwk{A}{B}}}
  \jdeq
f.\qedhere
\end{equation*}
\end{proof}

\begin{rmk}
Since any E-homomorphism is both a substitution homomorphism and a weakening
homomorphism, it follows immediately that any E-homomorphism preserves any
$f_\ast$, i.e.~for any $f\in\thom{A}{B}$ in $\mathcal{E}$ and any E-homomorphism
$F:\mathcal{E}\to\mathcal{D}$, the diagram
\begin{equation*}
\begin{tikzcd}[column sep=large]
\mathcal{E}/\ctxext{\Gamma}{A} \arrow[r,"F/\ctxext{\Gamma}{A}"] & \mathcal{D}/F(\ctxext{\Gamma}{A})
  \\
\mathcal{E}/\ctxext{\Gamma}{B} \arrow[u,"f_\ast"] \arrow[r,swap,"F/\ctxext{\Gamma}{B}"] & \mathcal{D}/F(\ctxext{\Gamma}{B}) \arrow[u,swap,"F(f)_\ast"]
\end{tikzcd}
\end{equation*}
commutes. This applies to weakening, substitution and pre-composition, and it
follows that for any E-homomorphism $F:\mathcal{E}\to\mathcal{D}$ and any
$\Gamma\in\mathcal{E}$, we have a functor $\mathbb{C}_I(F,\Gamma):
\mathbb{C}_I(\mathcal{E},\Gamma)\to\mathbb{C}_I(\mathcal{D},F(\Gamma))$.
\end{rmk}

@@

\begin{lem}\label{lem:compW_W}
Let $f\in\thom{A}{B}$ be an internal morphism in context $\Gamma$. Then one has
\begin{equation*}
f_\ast\circ W_B\jdeq W_A.
\end{equation*}
\end{lem}

\begin{proof}
The proof is a simple calculation:
\begin{equation*}
f_\ast\circ W_B\jdeq S_f\circ W_A/B\circ W_B\jdeq S_f\circ W_{W_A(B)}\circ W_A
\jdeq W_A.\qedhere
\end{equation*}
\end{proof}

@@

\begin{thm}\label{precomp_by_proj}
Precomposition with $\cprojfstf{A}{P}$ is weakening by $P$, i.e.
\begin{equation*}
(\cprojfstf{A}{P})_\ast \jdeq W_P.
\end{equation*}
\end{thm}

\begin{proof}
\begin{align*}
(\cprojfstf{A}{P})_\ast & \jdeq S_{\cprojfstf{A}{P}}\circ W_{\ctxext{A}{P}} \\
& \jdeq S_{W_P(\tfid{A})}\circ W_P\circ W_A \\
& \jdeq W_P\circ S_{\tfid{A}}\circ W_A \\
& \jdeq W_P.\qedhere 
\end{align*}
\end{proof}

@@

\begin{defn}
Let $\jhom{\Gamma}{A}{B}{f}$ and $\jfhom{\Gamma}{A}{B}{f}{P}{Q}{F}$. Then we
define the E-homomorphism 
\begin{equation*}
F_\circledast\defeq F_\ast\circ (f_\ast/Q):
\mathbb{E}/\ctxext{{\Gamma}{B}}{Q}\to\mathbb{E}/\ctxext{{\Gamma}{A}}{P}.
\end{equation*}
The infix notation of $F_\circledast$ is taken to be $\jfcomp{\blank}{\blank}{\blank}{F}{\blank}$.
\end{defn}

\begin{lem}\label{lem:three-composition}
Let $\jhom{\Gamma}{A}{B}{f}$ and $\jfhom{\Gamma}{A}{B}{f}{P}{Q}{F}$. Then we
have the equality
\begin{equation*}
F_\circledast\jdeq (\jvcomp{P}{f}{F})_\ast.
\end{equation*}
\end{lem}

>>>>

\section{Elementary properties of telescopic type systems}\label{sec:esys_props}

The theory of telescopic type systems could be seen as a sort of category theory with only
projections and their sections. We elaborate on this point of view by defining
the pre-category of internal morphisms, which Lawvere might have called the
category of terms. 
In the pre-category of internal morphisms over a context $\Gamma$,
which we define in \autoref{subsec:im_cat},
a morphism $f:A\to B$ is a term of the constant family $W_A(B)$ over 
$\ctxext{\Gamma}{A}$. Rather than directly defining composition, we will define
for every $f:A\to B$ in context $\Gamma$, a pre-composition operation
$f^\ast$, which is going to be an E-homomorphism from 
$\cat{F}/\ctxext{\Gamma}{B}$ to $\cat{F}/\ctxext{\Gamma}{A}$. This allows us
to consider diagrams of the following form:
\begin{equation*}
\begin{tikzcd}
P \arrow[d,->>] \arrow[r,"g"] & Q \arrow[d,->>] \\
A \arrow[r,swap,"f"] & B
\end{tikzcd}
\end{equation*}
where we consider the morphism $g$ over $f$ to be an internal morphism from 
$P$ to $f^\ast(Q)$ in context $\ctxext{\Gamma}{A}$, capturing the
essence of being a morphism from the family $P$ over $A$ to the family
$Q$ over $B$, reindexed by $f$. 
Whenever we draw such diagrams, a double headed arrow will always
refer to a family, and an ordinary arrow will refer to an internal morphism,
possibly over another internal morphism.

@@

The pre-category of internal morphisms forms a pre-category of which the families form
a subpre-category. This, together with the pre-composition operation described
earlier, gives rise to the auxiliary notion of E'-systems, which we define
in \autoref{sec:E'}. The thread running through this whole section is proving
that the notions of telescopic type system and E'-system are the same. In \autoref{sec:E'}
we show directly that any E'-system is a telescopic type system. We end this section by
showing that the converse also holds.

@@

\begin{defn}\label{defn:E'sys}
An \define{E'-system} consists of a pre-category $\cat{C}$ and a subpre-category $\cat{F}$ with
the same objects as $\cat{C}$, such that
\begin{enumerate}
\item $\cat{F}$ has a terminal object.
\item For any $f:\Delta\to\Gamma$ in $\cat{C}$ and any $A\in\cat{F}/\Gamma$, a
functorial choice of a pullback square 
\begin{equation*}
\begin{tikzcd}[column sep=large]
\ctxext{\Delta}{f^\ast(A)} \arrow[r,"{\pi_2(f,A)}"] \arrow[d,fib,swap,"{f^\ast(A)}"] & \ctxext{\Gamma}{A} \arrow[d,fib,"A"] \\
\Delta \arrow[r,"f"] & \Gamma
\end{tikzcd}
\end{equation*}
such that $f^\ast(A)\in\cat{F}/\Delta$. Explicitly, the functoriality
requirement is that
\begin{enumerate}
\item For any $f:\Delta\to\Gamma$, one has 
$f^\ast(\catid{\Gamma})\jdeq\catid{\Delta}$ and
\begin{equation*}
\pi_2(f,\catid{\Gamma})\jdeq f.
\end{equation*}
\item \label{defn:E'sys:b} For any $A\in\cat{F}/\Gamma$, one has $(\catid{\Gamma})^\ast(A)\jdeq A$ and
\begin{equation*}
\pi_2(\catid{\Gamma},A)\jdeq\catid{\ctxext{\Gamma}{A}}
\end{equation*}
\item \label{defn:E'sys:c} For any $f:\Delta\to\Gamma$, $g:\greek{E}\to\Delta$ and $A\in\cat{F}/\Gamma$, one
has $(f\circ g)^\ast(A)\jdeq g^\ast(f^\ast(A))$ and 
\begin{equation*}
\pi_2(f\circ g,A)\jdeq \pi_2(f,A)\circ\pi_2(g,f^\ast(A))
\end{equation*}
\item For any $P\in\cat{F}/\ctxext{\Gamma}{A}$ and $f:\Delta\to\Gamma$, one has
$f^\ast(P)\jdeq\pi_2(f,A)^\ast(P)$ and 
\begin{equation*}
\pi_2(f,\ctxext{A}{P}) \jdeq \pi_2(\pi_2(f,A),P)
\end{equation*}
\end{enumerate}
\end{enumerate}
For any $f:\Delta\to\Gamma$ we shall write $f^\ast$ for the induced functor
$\cat{F}/\Gamma\to\cat{F}/\Delta$. 
\end{defn}

\begin{rmk}
Without a terminal object in $\cat{F}$, the pre-category $\cat{C}$ does not correspond
directly to anything in a telescopic type system.
\end{rmk}

@@

\begin{lem}\label{lem:pb_selfdistributive}
Let $\mathbb{E}$ be an E'-system. Then $f^\ast$ can be extended to a functor from
$\cat{C}_\cat{F}/\Gamma$ to $\cat{C}_\cat{F}/\Delta$, for each $f:\Delta\to\Gamma$, and moreover the diagram
\begin{equation*}
\begin{tikzcd}[column sep=large]
\cat{C}_\cat{F}/\ctxext{\Gamma}{B} \arrow[r,"{f^\ast/B}"] \arrow[d,swap,"{g^\ast}"]
& \cat{C}_\cat{F}/\ctxext{\Delta}{f^\ast(B)} \arrow[d,"{(f^\ast(g))^\ast}"] \\
\cat{C}_\cat{F}/\ctxext{\Gamma}{A} \arrow[r,swap,"{f^\ast/A}"]
& \cat{C}_\cat{F}/\ctxext{\Delta}{f^\ast(A)}
\end{tikzcd}
\end{equation*}
commutes for every $g:A\to B$ in $\cat{C}_\cat{F}/\Gamma$.
\end{lem}

\begin{proof}
It is straightforward to use the universal property of the pullback to extend
$f^\ast$ to a functor from $\cat{C}_\cat{F}/\Gamma$ to $\cat{C}_\cat{F}/\Delta$.
To see that the diagram commutes, we calculate
\begin{align*}
f^\ast/A\circ g^\ast
& \jdeq
\pi_2(f,A)^\ast\circ g^\ast \\
& \jdeq
(g\circ\pi_2(f,A))^\ast \\
& \jdeq
(\pi_2(f,B)\circ f^\ast(g))^\ast \\
& \jdeq
(f^\ast(g))^\ast\circ\pi_2(f,B)^\ast \\
& \jdeq
(f^\ast(g))^\ast\circ f^\ast/B.\qedhere
\end{align*}
\end{proof}

\begin{thm}
A telescopic type system is the same thing as an E'-system.
\end{thm}

\begin{proof}
We define for any $A\in\cat{F}/\Gamma$ the set
\begin{equation*}
T(A)\defeq\{x:\Gamma\to\ctxext{\Gamma}{A}\mid A\circ x\jdeq\catid{\Gamma}\}.
\end{equation*}
Since each functor $f^\ast$ is extended to all of $\cat{C}_{\cat{F}}/\Gamma$,
it follows immediately that those are functors with term structure.

We define for any $A\in\cat{F}/\Gamma$, the functor $W_A\defeq A^\ast:\cat{F}/\Gamma\to\cat{F}/\ctxext{\Gamma}{A}$.
Likewise, we define for any $x\in T(A)$, the functor $S_x\defeq x^\ast:\cat{F}/\ctxext{\Gamma}{A}\to\cat{F}/\Gamma$.
We also define $\tfid{A}:T(W_A(A))$ by the universal property of pullbacks:
\begin{equation*}
\begin{tikzcd}[column sep=large]
\ctxext{\Gamma}{A} \arrow[drr,bend left=15,equals] \arrow[ddr,bend right=15,equals] \arrow[dr,densely dotted,near end,"{\tfid{A}}"] \\
& \ctxext{{\Gamma}{A}}{W_A(A)} \arrow[r,swap,"{\pi_2(A,A)}"] \arrow[d,fib] & \ctxext{\Gamma}{A} \arrow[d,fib] \\
& \ctxext{\Gamma}{A} \arrow[r,swap,"A"] & \Gamma
\end{tikzcd}
\end{equation*}
As an immediate consequence of \autoref{lem:pb_selfdistributive}, we get that
each functor $W_A$ and $S_x$ is both a weakening functor and a substitution functor.
Since each $W_A$ and $S_x$ preserves identity morphisms, it also preserves the
terms $\tfid{A}$.

It remains to verify the rest of the conditions of telescopic type systems.
\begin{itemize}
\item To show that weakened families are constant, note that
\begin{equation*}
S_x\circ W_A\jdeq x^\ast\circ A^\ast\jdeq (A\circ x)^\ast\jdeq (\catid{\Gamma})^\ast\jdeq \catid{\cat{C}_{\cat{F}}/\Gamma}.
\end{equation*}
\item The identity terms are neutral for pre-composition:
\begin{equation*}
S_{\tfid{A}}\circ W_A/A \jdeq S_{\tfid{A}}\circ \pi_2(A,A)^\ast \jdeq (\pi_2(A,A)\circ\tfid{A})^\ast \jdeq
(\catid{\ctxext{\Gamma}{A}})^\ast \jdeq \catid{\cat{C}_{\cat{F}}/\ctxext{\Gamma}{A}} 
\end{equation*}
\item The identity terms behave like identity functions: by the universal property,
$S_x(\tfid{A})$ is the unique section of $A$ such that the square
\begin{equation*}
\begin{tikzcd}[column sep=6em]
\Gamma \arrow[d,swap,"{S_x(\tfid{A})}"] \arrow[r,"{\pi_2(x,\catid{\ctxext{\Gamma}{A}})}"] & \ctxext{\Gamma}{A} \arrow[d,"{\tfid{A}}"] \\
\ctxext{\Gamma}{A} \arrow[r,swap,"{\pi_2(x,W_A(A))}"] & \ctxext{{\Gamma}{A}}{W_A(A)}
\end{tikzcd}
\end{equation*}
commutes. Thus, it suffices to show that this square also commutes with $x$ in the place of
$S_x(\tfid{A})$. Note that $\pi_2(x,\catid{\ctxext{\Gamma}{A}})\jdeq x$. 
Since $\ctxext{{\Gamma}{A}}{W_A(A)}$ is itself a pullback, it suffices
and it is straightforward to verify the equalities
\begin{align*}
W_A(A)\circ\pi_2(x,W_A(A))\circ x & \jdeq W_A(A)\circ\idtm{A}\circ x \\
\pi_2(A,A)\circ\pi_2(x,W_A(A))\circ x & \jdeq \pi_2(A,A)\circ\idtm{A}\circ x.\qedhere
\end{align*}
\end{itemize}
\end{proof}

\begin{comment}
We then demonstrate how to define a category internal to a telescopic type system. The main
point of this lies mainly in demonstrating how the operations of telescopic type systems can
be used to define a category, because it is no surprise that a category can be
defined in type theory. Nevertheless, it is noteworthy that we need no further 
conditions on telescopic type systems.
This is unlike the situation in category theory, where we need
to assume that the ambient category has finite limits in order to say what an
internal category is. In fact, it is possible to define a telescopic type system internal
to a telescopic type system, just using the algebraic structure of a telescopic type system, and actually
defining an internal category was the hardest part of that definition.
\end{comment}

@@

\begin{defn}
Let $A,B\in\cat{F}/\Gamma$.
For any $f\in\thom{A}{B}$ we define the \define{pre-composition} E-homomorphism
\begin{equation*}
f^\ast \defeq S_f\circ W_A/B : \cat{F}/\ctxext{\Gamma}{B}\to\cat{F}/\ctxext{\Gamma}{A}.
\end{equation*}
Furthermore, for $g\in\thom{B}{C}$ we may consider $W_B(C)\in\cat{F}/\ctxext{\Gamma}{B}$ as a morphism from $W_B(C)$ to $\catid{\ctxext{\Gamma}{B}}$. Therefore, $f^\ast$ acts on the terms of
$W_B(C)$, so it makes sense to define $g\circ f\defeq f^\ast(g)$. In fact, it makes
sense to let $(\blank)\circ f$ be the infix notation for $f^\ast$. 
\end{defn}

@@

\begin{rmk}
Note that \autoref{f:SidfWff_id} of \autoref{defn:esystems} asserts precisely
that ${(\tfid{A})}^\ast\jdeq \catid{\cat{F}/\ctxext{\Gamma}{A}}$ for any $A\in\cat{F}/\Gamma$. In
particular, it follows that $g\circ\tfid{A}\jdeq g$ for any $g\in\thom{A}{B}$
\end{rmk}

\begin{lem}\label{lem:compcomp}
For any $f\in\thom{A}{B}$ and $g\in\thom{B}{C}$ we have $f^\ast\circ g^\ast\jdeq (g\circ f)^\ast$. 
\end{lem}

\begin{proof}
\begin{align*}
f^\ast\circ g^\ast & \jdeq S_f\circ (W_A/B) \circ S_g \circ (W_B/C)\\
& \jdeq S_f\circ S_{W_A(g)}\circ(W_A/\ctxext{B}{W_B(C)})\circ (W_B/C) \\
& \jdeq S_{S_f(W_A(g))}\circ (S_f/W_A(W_B(C)))\circ (W_A/\ctxext{B}{W_B(C)}) \circ W_B/C \\
& \jdeq S_{S_f(W_A(g))}\circ ((S_f\circ (W_A/B) \circ W_B)/C) \\
& \jdeq S_{S_f(W_A(g))}\circ ((S_f\circ W_{W_A(B)}\circ W_A)/C) \\
& \jdeq S_{S_f(W_A(g))}\circ W_A/C \\
& \jdeq (g\circ f)^\ast.\qedhere
\end{align*}
\end{proof}

\begin{thm}
For each object $\Gamma$ of a telescopic type system $\mathbb{E}$, we have a pre-category
$\mathbb{C}_I(\mathbb{E},\Gamma)$ of internal morphisms of $\mathbb{E}$,
consisting of the same objects as $\cat{F}/\Gamma$ and the hom-sets 
$\thom{A}{B}$. 
\end{thm}

\begin{proof}
The fact that composition is associative is a direct corollary of
\autoref{lem:compcomp}. The axiom $(\tfid{A})^\ast\jdeq\catid{\ctxext{\Gamma}{A}}$
implies that the identity morphisms satisfy the right identity law. It remains
to show that $\tfid{B}\circ f\jdeq f$. This is a simple calculation:
\begin{equation*}
\tfid{B}\circ f 
  \jdeq
\subst{f}{\ctxwk{A}{\tfid{B}}}
  \jdeq
\subst{f}{\tfid{\ctxwk{A}{B}}}
  \jdeq
f.\qedhere
\end{equation*}
\end{proof}

\begin{rmk}
Since any E-homomorphism is both a substitution homomorphism and a weakening
homomorphism, it follows immediately that any E-homomorphism preserves any
$f^\ast$, i.e.~for any $f\in\thom{A}{B}$ in $\mathcal{E}$ and any E-homomorphism
$F:\mathcal{E}\to\mathcal{D}$, the diagram
\begin{equation*}
\begin{tikzcd}[column sep=large]
\mathcal{E}/\ctxext{\Gamma}{A} \arrow[r,"F/\ctxext{\Gamma}{A}"] & \mathcal{D}/F(\ctxext{\Gamma}{A})
  \\
\mathcal{E}/\ctxext{\Gamma}{B} \arrow[u,"f^\ast"] \arrow[r,swap,"F/\ctxext{\Gamma}{B}"] & \mathcal{D}/F(\ctxext{\Gamma}{B}) \arrow[u,swap,"F(f)^\ast"]
\end{tikzcd}
\end{equation*}
commutes. This applies to weakening, substitution and pre-composition, and it
follows that for any E-homomorphism $F:\mathcal{E}\to\mathcal{D}$ and any
$\Gamma\in\mathcal{E}$, we have a functor $\mathbb{C}_I(F,\Gamma):
\mathbb{C}_I(\mathcal{E},\Gamma)\to\mathbb{C}_I(\mathcal{D},F(\Gamma))$.
\end{rmk}

@@

\begin{lem}\label{lem:compW_W}
Let $f\in\thom{A}{B}$ be an internal morphism in context $\Gamma$. Then one has
\begin{equation*}
f^\ast\circ W_B\jdeq W_A.
\end{equation*}
\end{lem}

\begin{proof}
The proof is a simple calculation:
\begin{equation*}
f^\ast\circ W_B\jdeq S_f\circ W_A/B\circ W_B\jdeq S_f\circ W_{W_A(B)}\circ W_A
\jdeq W_A.\qedhere
\end{equation*}
\end{proof}

@@

\begin{thm}\label{precomp_by_proj}
Precomposition with $\cprojfstf{A}{P}$ is weakening by $P$, i.e.
\begin{equation*}
(\cprojfstf{A}{P})^\ast \jdeq W_P.
\end{equation*}
\end{thm}

\begin{proof}
\begin{align*}
(\cprojfstf{A}{P})^\ast & \jdeq S_{\cprojfstf{A}{P}}\circ W_{\ctxext{A}{P}} \\
& \jdeq S_{W_P(\tfid{A})}\circ W_P\circ W_A \\
& \jdeq W_P\circ S_{\tfid{A}}\circ W_A \\
& \jdeq W_P.\qedhere 
\end{align*}
\end{proof}

@@

\begin{defn}
Let $\jhom{\Gamma}{A}{B}{f}$ and $\jfhom{\Gamma}{A}{B}{f}{P}{Q}{F}$. Then we
define the E-homomorphism 
\begin{equation*}
F^\circledast\defeq F^\ast\circ (f^\ast/Q):
\mathbb{E}/\ctxext{{\Gamma}{B}}{Q}\to\mathbb{E}/\ctxext{{\Gamma}{A}}{P}.
\end{equation*}
The infix notation of $F^\circledast$ is taken to be $\jfcomp{\blank}{\blank}{\blank}{F}{\blank}$.
\end{defn}

\begin{lem}\label{lem:three-composition}
Let $\jhom{\Gamma}{A}{B}{f}$ and $\jfhom{\Gamma}{A}{B}{f}{P}{Q}{F}$. Then we
have the equality
\begin{equation*}
F^\circledast\jdeq (\jvcomp{P}{f}{F})^\ast.
\end{equation*}
\end{lem}
